DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-11-17 2015CompleteDerivation/completeDerivation.tex
Reviewed first to subsections for quantum systems

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we now explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space of a composite system and the state space of its parts. In general, this is quite a complicated thing to do, which requires intimate knowledge of the system at hand. So we simplify our problem and study a system made of infinitesimal homogeneous parts. What we'll find is that under the additional assumption that the system is infinitesimally reducible (i.e. its state is equivalent to the states of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian mechanics.

@@

In this section we explore a different relationship between the whole material and its parts from the one described in section \ref{sec:reducibility}. The system is still infinitesimally decomposable, made of infinitesimal parts as before, but this time it is irreducible. That is: the state and the evolution of the whole does not tell us anything about the state and the evolution of the parts. As the internal dynamics is undetermined, we'll find that each irreducible element is suitably described by the standard framework of quantum mechanics.

@@

The first possible source of confusion is between divisibility, the idea that an object can be divided into parts, and reducibility, the idea that the state of the whole is equivalent to the state of the parts. Most of the time these two concepts overlap, but they are separate as we can see in the following examples.

@@

Consider a muon. It will decay into an electron and two neutrinos. But the state of a muon is not the state of an electron and two neutrinos. The muon is divisible into the three particles but it is not reducible into them. Consider a proton. Its state is given by the state of its three valence quarks. Yet, we cannot divide the proton into the three quarks. The proton is reducible to quarks, but divisible into quarks.

@@

Even more subtle is the difference between divisibility and decomposability, the idea that the object has parts. While we are agnostic on whether a quantum particle can be divided, we are implicitly assuming that it can be decomposed. Irreducibility describes the relationship between the state of the whole and the state of the parts, therefore parts are implicit in the definition. While these parts must exist, we cannot isolate them and act on them independently: if we were able to do that we could assign them a state.

While we must be able to talk about parts of a quantum particles, not all questions will make sense. Can we ask what fraction of a quantum particle is in a particular region of space? Yes, since the particle is decomposed some part may be inside the volume. Can we ask where that same fraction will be at a later time? No, since the particle is irreducible we cannot describe the evolution of a part.

The proper way to think of a quantized material is therefore the following: a continuum of infinitesimal parts clumped together to form quantum particles, the smallest amount of material to which we can assign a state. With this picture in mind, we can see, even before developing the math, how some features typical of quantum behaviors emerge.

The first feature is the importance of statistics in the quantum particle state. Consider a part of a quantum particle in an initial state. As we cannot track its motion, it can be any part of the particle in its final state. That is, the motion of a part of a particle is a random walk within the evolution of a particle.

The second feature is non locality. Suppose we have a quantum particle state distributed in space. Suppose we exert a force only on a subregion of its distribution. As the particle is irreducible, we cannot interact with only part of it: we affect also the part where no force is exerted. We cannot simply interact with the part at a specific region, as this would allow to assign a state to that part.

Wave particle duality should be clear at this point. The quantum particle behaves like a wave in the sense that its parts form a distribution in space. Yet, it behaves as a single unit, as the state is one unit and cannot be separated.

The most striking feature, though, is that we can predict the functional form for particle decay. Suppose a quantum particle spontaneously decays into two or more particles. Such process cannot be deterministic as the d.o.f. increase, therefore is not determined by the state of the quantum particle. As the decay is spontaneous, it is not triggered by the environment (i.e. external state). It is therefore triggered by the internal unstated dynamic of the parts. As we assumed the internal dynamic is unknown, we cannot learn anything about it by looking at the product of the decay. In particular, all states have to have equal chance of decay or we could tell that the internal dynamic is more stable for some states. The probability of decay is constant in time. The probability of survival follows the exponential distribution.

@@

\subsection{Quantum particle and complex vector spaces}

Turning these concept into a precise mathematical formulation is tricky as we need to formally characterize the unstated part which is not physically distinguishable. While many previous results do not apply, the classical derivation still can serve as a template.

As in the classical case, the quantum particle is thought to be decomposable: different parts will be in different places, the evolution of the whole is the composition of the evolution of parts. Therefore \ref{prop:abelian_group} applies and the state space $\Psi$ of an irreducible material is an abelian group.

Note, though, that we can't think of the parts as independent components. As the unstated part is statistical in nature, it will combine as statistical variables do. Suppose $X$ and $Y$ are two random variables, $\sigma^2_{X+Y} = \sigma^2_{X} + \sigma^2_{Y} + 2 \, \sigma_{X} \sigma_{Y} \rho_{X,Y}$, where $\sigma^2$ is the variance, $\sigma$ is the standard deviation, and $\rho_{X,Y}=COV(X,Y)/\sigma_{X} \sigma_{Y}$ is the Pearson correlation coefficient. As $-1\leq\rho_{X,Y}\leq+1$, the total variance can be greater or smaller then the sum of the independent parts, and the strength of the uncertainty within the unstated part is something physically distinguishable.

Therefore, while the motion of the unstated part is not itself physically relevant, the correlation between the motion of the unstated part of two combined parts is very relevant. We'll see how interference can be understood in this context: it is constructive if the randomness is correlated while it is destructive if it is anti-correlated.

So, how can we characterized these relationship between different configurations of the unstated parts? We can look at the classical derivation for guidance. In that case, we looked at the set of transformation that increase/decrease the amount of material. Those are the only type of transformation that we can apply on a classical system that do not change the overall state given by the distribution.

As the classical material is homogeneous, permutations between classical particles in the same state $\mathcal{s}$ do not matter: the ordering is unstated. In a way, increasing/decreasing the amount of material increases/decreases the unstated part of the system.

For the irreducible material we follow the same strategy: we find the state transformations that affect the unstated part while keeping the overall state, the distribution, unchanged.

Suppose we have $\psi_1$ a state for an irreducible material characterized by one d.o.f. The material will be spread over the configurations of the infinitesimal fragments. As the material is homogeneous, the material is spread evenly among the such possible configuration. Counting configurations is therefore equivalent to counting amount of material. Each fragment would be identified by a pair of independent random variables $(A, B)$, as a d.o.f. is two-dimensional. The shape of the distribution and the actual values are not physically distinguishable: these are just dummy variables we assign to count the fragment configurations. We choose A and B to be uniform distributions from 0 to 1, so that the area is a square of unit size, which represent the unit of unstated  configurations the material is spread over.

Suppose we have another state $\psi_2$, where only unstated part is changed. The unstated part can also be characterized by a square formed by two independent uniform distributions (C, D). Counting states on (C,D) is equivalent to counting them on (A,B), as they are simply different arrangements. As all distributions are uniform, they can only be connected by a linear operation. That is: $C=aA+bB$ and $D= -bA + aB$. We can write $\psi_2 = \tau(a,b) \psi_1$ as $a$ and $b$ are the only parameters of the transformation.

The transformation $\tau(a,b)$ increases/decreases the area by the factor $a^2+b^2$ and rotates the square by $arctan(x/y)$. If we set $c=a+\imath b$ those same quantities can be expressed as $\rho^2=|c|^2$ and $\theta = arg(z)$, the modulus and the phase of a complex number. In other words: complex numbers provide a representation for the group of rotations and magnitude rescaling.

For multiple independent d.o.f., we have one square each. As the material is irreducible, all areas need to be always the same since relative changes would allow to learn something about the motion of the infinitesimal fragments. This means that the transformation $\tau(c)$ has to be the same for all d.o.f. In other words, the group of transformations is the same regardless of the number of independent d.o.f.

Mathematically, the state space $\Psi$ for an irreducible material is a complex vector space.

@@

>>>>

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we now explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space of a composite system and the state space of its parts. In general, this is quite a complicated thing to do, which requires intimate knowledge of the system at hand. So we simplify our problem and study a material made of infinitesimal homogeneous parts. What we'll find is that under the additional assumption that the material is infinitesimally reducible (i.e. its state is equivalent to the states of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian mechanics.

@@

In this section we explore a different relationship between the whole and its parts. We will consider an irreducible material, one where the state and the evolution of the whole does not tell us anything about the state and the evolution of the parts. The material is still conceptually infinitesimally decomposable, made of infinitesimal parts which we call fragments. But each fragment cannot be studied independently and therefore cannot be assigned a state: the internal dynamics is undetermined. We'll find that the evolution of an irreducible material is suitably described by the standard framework of quantum mechanics.

@@

The first possible source of confusion is between divisibility, the idea that an object can be divided into parts, and reducibility, the idea that the state of the whole is equivalent to the state of the parts. More precisely, a system is divisible if there is a time evolution $\mathcal{T}_{\Delta t}: \mathcal{S} \rightarrow \mathcal{S}_1 \times \mathcal{S}_2$ that takes the state of a system and returns the state of two separate isolated systems. It is reducible if $\mathcal{S} = \mathcal{S}_1 \times \mathcal{S}_2$, the state of the system is the state of the two subsystems. Most of the time these two concepts overlap, but they are separate as we can see in the following examples.

@@

Consider a muon. It will decay into an electron and two neutrinos. But the state of a muon is not the state of an electron and two neutrinos. The muon is divisible into the three particles but it is not reducible to them. Consider a proton. Its state is given by the state of its three valence quarks and the gluon field. Yet, we cannot divide the proton into the three isolated quarks. The proton is reducible to quarks, but not divisible into quarks.

@@

Even more subtle is the difference between divisibility and decomposability, the idea that the object has parts. That is, a system is decomposable if there is a law $+ : \mathcal{S} \times \mathcal{S} \rightarrow \mathcal{S}$ to combine two states to get a third. While irreducibility says nothing on whether a quantum particle can be divided, it implicitly assumes that it can be decomposed. Irreducibility describes the relationship between the state of the whole and the state of the parts, therefore the existence of parts are implicit in the definition. But it also tells us that we are limited in what description we can give to these parts: we cannot isolate them as we want and act on them independently because if we were able to do that we could assign them an independent state.

Therefore, while we must be able to talk about parts of a quantum particles, not all questions will make sense. Can we ask what fraction of a quantum particle is in a particular region of space? Yes, since the particle can be spatially decomposed some amount of material may be inside the volume. Can we ask where that same fraction will be at a later time? No, since the particle is irreducible we cannot describe the evolution of a part.

Note that if the material is infinitesimally reducible, reducibility/decomposability/divisibility are all possible. This is why one may not dwell on their difference and use the word ``part" interchangeably in the three cases. Now that we assume irreducibility, it is better to use more precise language to avoid confusion. Therefore we will call \emph{fragments} the infinitesimal parts of the material that cannot be described independently. We will call \emph{components} the parts that the material can be decomposed into. A component will have a state associated with it, and therefore will always be a collection of fragments. As in the classical case, a component may be the difference between two states, therefore one should not imagine it always ``contained" in the composed system.

The proper way to think of a quantized material is therefore the following: a continuum of infinitesimal fragments clumped together to form quantum particles, the smallest amount of material to which we can assign a state. With this picture in mind, we can see, even before developing the math, how some features typical of quantum behaviors emerge.

The first feature is the importance of statistics. Consider a fragment of a quantum particle in an initial state. As we cannot track its motion, it can correspond to any fragment of the particle in its final state. That is, the motion of a fragment of a particle is a random walk within the evolution of a particle.

The second feature is non locality. Suppose we have a quantum particle distributed in space. Suppose we exert a force only on a subregion of its distribution. As the particle is irreducible, we cannot interact with only part of it: we must also affect the part where no force is exerted. We cannot simply interact with the part at a specific region, as this would allow to assign a state to that part.

Wave particle duality is also accounted for. The quantum particle behaves like a wave in the sense that its parts form a distribution in space. Yet, it behaves as a single unit, as the state is irreducible and cannot be split.

The most striking feature, though, is that we can predict the functional form for particle decay over time. Suppose a quantum particle spontaneously decays into two or more particles. Such process cannot be deterministic as the number of d.o.f. increases. As the decay is spontaneous, it is not triggered by the environment (i.e. external state). It must be triggered by the internal dynamic of the fragments. As we assumed the internal dynamic is unknown, we cannot learn anything about it by looking at the product of the decay. In particular, all states have to have equal chance of decay or we could tell that the internal dynamic is more stable for some states. The probability of decay is constant for all states at all times. The probability of survival follows the exponential distribution.

@@

\subsection{Irreducible material and complex vector spaces}

It is time to turn these concepts into a precise mathematical formulation following the template used for classical mechanics. We want to formalize the state space for an \emph{irreducible material} as a quantum particle will be a finite amount of such material.

As for the classical material, the irreducible material is thought to be decomposable: each state can seen as the sum of different components (e.g. spatially distributed part) such that the evolution of the whole is the composition of their evolution. Therefore \ref{prop:abelian_group} applies and the state space $\Psi$ of an irreducible material is an abelian group.

Note, though, that the components are in general not independent. As the motion of each fragment is statistical in nature, it will combine as statistical variables do. Suppose $X$ and $Y$ are two random variables,
\begin{equation}\label{eqn:correlation}
\sigma^2_{X+Y} = \sigma^2_{X} + \sigma^2_{Y} + 2 \, \sigma_{X} \sigma_{Y} \rho_{X,Y}
\end{equation}
where $\sigma^2$ is the variance, $\sigma$ is the standard deviation, and $\rho_{X,Y}=COV(X,Y)/\sigma_{X} \sigma_{Y}$ is the Pearson correlation coefficient. As $-1\leq\rho_{X,Y}\leq+1$, the total variance can be greater or smaller then the sum of of the variance of the components. The strength of the chaotic motion of the fragments is something physically distinguishable as it contributes to the spread of the distribution. Therefore while the motion of each fragment is not itself physically distinguishable, it falls within the unstated part, the strength and correlation of the chaotic motion must be, they fall within the state.

How can we characterize the aspects of the unstated part that are captured by the state? We can look at the classical derivation for guidance. In that case, we looked at the set of transformation that increase/decrease the amount of material. We can think them as the transformations that change the unstated part of the system. As the classical material is homogeneous, permutations between classical particles in the same state $\mathcal{s}$ do not matter (i.e. the ordering is unstated). Increasing/decreasing the amount of material increases/decreases the number of permutations and therefore changes the unstated part of the system. For the irreducible material we follow the same strategy: we look for the transformations that affect the unstated part while keeping the overall state, the distribution, unchanged.

Suppose we have $\psi_1$ a state for an amount $a_m$ of irreducible material characterized by one d.o.f. The material will be spread over the configurations of the infinitesimal fragments. As the material is homogeneous, the material is spread evenly among such possible configurations. Counting fragment configurations is therefore equivalent to counting amount of material. Each fragment would be identified by a pair of independent random variables $(A, B)$, as a d.o.f. is two-dimensional. The shape of the distribution and the actual values are not physically distinguishable: these are just dummy variables we assign to count the fragments. We can choose A and B to be uniform distributions of interval $\sqrt{a_m}$, so that they identify a square of area $a_m$, the amount of material.

Suppose we have another configuration $\psi_2$, where only the unstated part is changed. The unstated part of $\psi_2$ can also be characterized by a square formed by two independent uniform distributions (C, D). Counting states on (C,D) is equivalent to counting them on (A,B), as they are simply different arrangements of fragments within the same state. The relationship between the two pairs of variables must be linear since all distributions are uniform. That is: $C=aA+bB$ and $D= -bA + aB$. We can write $\psi_2 = \tau(a,b) \psi_1$ as $a$ and $b$ are the only parameters of the transformation. Or more compactly, $\psi_2 = \tau(c) \psi_1$ where $c=a+\imath b$.

The transformation $\tau(c)$ increases/decreases the area by the modulus $|c|^2=a^2+b^2$, which physically corresponds to the increase/decrease of amount of material. It also rotates the square by the phase $\theta_c \equiv arg(c) = arctan(a/b)$, which corresponds in a change in correlation: if $\theta_c = 0$, $C = |c| A$ are clearly correlated while if $\theta_c = \pi$, $C = - |c| A$ are clearly anti-correlated.

To better understand the link between phase and correlation, suppose $\psi$ is the state for an irreducible material. We can write $\tau(c) \psi = \tau(c_1 + c_2) \psi = \tau(c_1) \psi + \tau(c_2) \psi$ as the composition of two separate transformations. We have
\begin{equation}\label{eqn:correlation_state}
|c|^2=|c_1|^2 + |c_2|^2 + 2 |c_1||c_2|\cos(\theta_{c_2} - \theta_{c_1})
\end{equation}
Compare to \ref{eqn:correlation}. The material is spread evenly among fragment configurations. Also, since it is homogeneous, the ratio of fragment per configuration is the same. Therefore the amount of material is proportional to the strength of the chaotic motion of the fragment, which is measured by the variance. The increase/decrease in amount of material therefore combines like the variance of a statistical variable. The cosine of the phase difference takes the place of the Pearson correlation coefficient. A more meaningful name for the phase, then, would be \emph{correlation angle}. Also note how correlation for a single variable is a meaningless quantity: it is the correlation between two random variables that is significant. That is why in quantum mechanics phase differences have physical meaning, as it gives the correlation between two states, while absolute phases do not.

If we increase the number of independent d.o.f. the group of transformations remains the same.  As the material is homogeneous, the same amount of material needs to be spread over the same number of fragment configurations along each independent d.o.f. This means that, as we combine components, the strength of the chaotic motion has to be the same in all directions. This means that the increase/decrease of material and the change in correlation has to be the same across all d.o.f.

Mathematically, the state space $\Psi$ for an irreducible material is a complex vector space. It is so because it is the state space for a decomposable material (i.e. an abelian group) where we can increase/decrease the amount and change the correlation of the internal stochastic motion (i.e. a set of transformation conveniently parametrized by a complex number).

@@

\subsection{Complex valued distributions}
